---
title: "Business Intelligence Lab Submission Markdown"
author: "Lumin"
date: "04/11/23"
output:
  github_document:
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
always_allow_html: true
editor_options:
  chunk_output_type: console
---


# Student Details

+---------------------------------------------------+------------------------------------------------------------------------------------------------------+
| **Student ID Numbers and Names of Group Members** | 
|                                                   |                                                                                                      |
|                                                   | 1.  112827 - A - Mungai Kenneth                                                                      |
|                                                   |                                                                                                      |
|                                                   | 2.  123324 - B - Kelly Noella Sota                                                                     |
|                                                   |                                                                                                      |
|                                                   | 3.  134265 - A - Emmanuel Kiptoo                                                                   |
|                                                   |                                                                                                      |
|                                                   |                                                                               |
|                                                   |                                                                                                      |
|                                                   |                                                                               |
+---------------------------------------------------+------------------------------------------------------------------------------------------------------+
| **GitHub Classroom Group Name**                   |    Lumin                                                                                                  |
+---------------------------------------------------+------------------------------------------------------------------------------------------------------+
| **Course Code**                                   | BBT4206                                                                                              |
+---------------------------------------------------+------------------------------------------------------------------------------------------------------+
| **Course Name**                                   | Business Intelligence II                                                                             |
+---------------------------------------------------+------------------------------------------------------------------------------------------------------+
| **Program**                                       | Bachelor of Business Information Technology                                                          |
+---------------------------------------------------+------------------------------------------------------------------------------------------------------+
| **Semester Duration**                             | 21^st^ August 2023 to 28^th^ November 2023                                                           |
+---------------------------------------------------+------------------------------------------------------------------------------------------------------+

# Setup Chunk

**Note:** the following "*KnitR*" options have been set as the defaults:\
`knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy.opts = list(width.cutoff = 80), tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
if (require("formatR")) {
  require("formatR")
} else {
  install.packages("formatR", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}
library(formatR)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE,
                      collapse = FALSE, tidy = TRUE)
```

**Note:** the following "*R Markdown*" options have been set as the defaults:

> output:\
> \
> github_document:\
> toc: yes\
> toc_depth: 4\
> fig_width: 6\
> fig_height: 4\
> df_print: default\
> \
> editor_options:\
> chunk_output_type: console


# 1: Install and Load Required Packages
In this step, we ensure that the necessary R packages are installed and loaded. Packages are collections of R functions, data, and compiled code that extend the functionality of R. The install.packages() function is used to install packages, and library() is used to load them.
```{r setup-chunk-one}
if (require("languageserver")) {
  require("languageserver")
} else {
  install.packages("languageserver", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

# STEP 1. Install and Load the Required Packages ----
## readr ----
if (require("readr")) {
  require("readr")
} else {
  install.packages("readr", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## naniar ----
if (require("naniar")) {
  require("naniar")
} else {
  install.packages("naniar", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## ggplot2 ----
if (require("ggplot2")) {
  require("ggplot2")
} else {
  install.packages("ggplot2", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## corrplot ----
if (require("corrplot")) {
  require("corrplot")
} else {
  install.packages("corrplot", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## ggcorrplot ----
if (require("ggcorrplot")) {
  require("ggcorrplot")
} else {
  install.packages("ggcorrplot", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## mlbench ----
if (require("mlbench")) {
  require("mlbench")
} else {
  install.packages("mlbench", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## stringi ----
if (require("stringi")) {
  require("stringi")
} else {
  install.packages("stringi", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## caret ----
if (require("caret")) {
  require("caret")
} else {
  install.packages("caret", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

## dplyr ----
if (require("dplyr")) {
  require("dplyr")
} else {
  install.packages("dplyr", dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

```

# 2. Load the Dataset
This code snippet is used to load the dataset "Air Traffic Passenger Statistics".
```{r step-two-chunk}
library(readr)

# Load the Air Traffic Passenger dataset
Air_Traffic_Passenger_Statistics <- read_csv("data/Air_Traffic_Passenger_Statistics.csv",
                                             col_types = cols(
                                               Activity_Period = col_integer(),
                                               Year = col_integer(),
                                               Passenger_Count = col_integer(),
                                               Adjusted_Passenger_Count = col_integer(),
                                               Month = col_character(),
                                               Boarding_Area = col_character(),
                                               Terminal = col_character(),
                                               Activity_Type_Code = col_character(),
                                               Price_Category_Code = col_character(),
                                               Published_Airline_IATA_Code = col_character(),
                                               GEO_Summary = col_character(),
                                               GEO_Region = col_character(),
                                               Operating_Airline = col_character(),
                                               Operating_Airline_IATA_Code = col_character(),
                                               Published_Airline = col_character(),
                                               Adjusted_Activity_Type_Code = col_character()
                                             )
)




str(Air_Traffic_Passenger_Statistics)
dim(Air_Traffic_Passenger_Statistics)
head(Air_Traffic_Passenger_Statistics)
summary(Air_Traffic_Passenger_Statistics)


```

# 3.  Check for Missing Data and Address it
This code checks if there are any missing values and try to correct it by removing them from the datasets. This code will showcase two options to deal with missing values. 
```{r step-three-chunk}
# Are there missing values in the dataset?
any_na(Air_Traffic_Passenger_Statistics)

# How many?
n_miss(Air_Traffic_Passenger_Statistics)

# What is the proportion of missing data in the entire dataset?
prop_miss(Air_Traffic_Passenger_Statistics)

# What is the number and percentage of missing values grouped by
# each variable?
miss_var_summary(Air_Traffic_Passenger_Statistics)

# Which variables contain the most missing values?
gg_miss_var(Air_Traffic_Passenger_Statistics)

# Which combinations of variables are missing together?
gg_miss_upset(Air_Traffic_Passenger_Statistics)

# Where are missing values located (the shaded regions in the plot)?
vis_miss(Air_Traffic_Passenger_Statistics) +
  theme(axis.text.x = element_text(angle = 80))

## OPTION 1: Remove the observations with missing values ----
# We can decide to remove all the observations that have missing values
# as follows:
AirTraffic_removed_obs <- na.omit(Air_Traffic_Passenger_Statistics)

# The initial dataset had 15007 observations and 16 variables
dim(Air_Traffic_Passenger_Statistics)

# The filtered dataset has 14953 observations and 16 variables
dim(AirTraffic_removed_obs)

## OPTION 2: Remove the variables with missing values ----
# Alternatively, we can decide to remove the 2 variables that have missing data
AirTraffic_removed_vars <-
  Air_Traffic_Passenger_Statistics %>%
  dplyr::select(- Operating_Airline_IATA_Code, - Published_Airline_IATA_Code)

# The initial dataset had 15007 observations and 16 variables
dim(Air_Traffic_Passenger_Statistics)

# The filtered dataset has 15007 observations and 14 variables
dim(AirTraffic_removed_vars)

# Are there missing values in the dataset?
any_na(AirTraffic_removed_vars)


```

# 4.Perform EDA and Feature Selection 
This code will compute the correlations between variables. Numeric variables are the ones that are going to be used. The goal is to find correlated variables so that they can be used to identify clusters.
```{r step-four-chunk}
# Create a correlation matrix
# Option 1: Basic Table
cor(AirTraffic_removed_obs[, c(1,12,14,15)]) %>%
  View()

# Option 2: Basic Plot
cor(AirTraffic_removed_obs[, c(1,12,14,15)]) %>%
  corrplot(method = "square")

# Option 3: Fancy Plot using ggplot2
corr_matrix <- cor(AirTraffic_removed_obs[, c(1,12,14,15)])

p <- ggplot2::ggplot(data = reshape2::melt(corr_matrix),
                     ggplot2::aes(Var1, Var2, fill = value)) +
  ggplot2::geom_tile() +
  ggplot2::geom_text(ggplot2::aes(label = label_wrap(label, width = 10)),
                     size = 4) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

ggcorrplot(corr_matrix, hc.order = TRUE, type = "lower", lab = TRUE)

#The correlation plot shows a 0.06 correlation between the Year and the
# Passenger Count.

# Other non-numeric variables and categorical variables are 
# not included in the correlation, but they can be used as an 
# additional dimension when plotting the scatter plot during EDA.


## Plot the scatter plots ----
# A scatter plot to show the activity period against year
# per terminal
ggplot(AirTraffic_removed_obs,
       aes(Passenger_Count, Year,
           color = Terminal)) +
  geom_point(alpha = 0.5) +
  xlab("Number of Passengers") +
  ylab("Number of Passengers/Terminal")

# A scatter plot to show the number of passenger against year
# per boarding area
ggplot(AirTraffic_removed_obs,
       aes(Passenger_Count, Year,
           color = Boarding_Area)) +
  geom_point(alpha = 0.5) +
  xlab("Number of Passengers") +
  ylab("Number of Passenger/Boarding Area")

# A scatter plot to show the number of passenger against year
# per geographic area
ggplot(AirTraffic_removed_obs,
       aes(Passenger_Count, Year,
           color = GEO_Region)) +
  geom_point(alpha = 0.5) +
  xlab("Number of Passengers") +
  ylab("Number of Passengers/GEO Region")


## Transform the data ----
# The K Means Clustering algorithm performs better when data transformation has
# been applied. This helps to standardize the data making it easier to compare
# multiple variables.

summary(AirTraffic_removed_obs)
model_of_the_transform <- preProcess(AirTraffic_removed_obs,
                                     method = c("scale", "center"))
print(model_of_the_transform)
AirTraffic_removed_obs_std <- predict(model_of_the_transform, # nolint
                                      AirTraffic_removed_obs)
summary(AirTraffic_removed_obs_std)
sapply(AirTraffic_removed_obs_std[, c(1,12,14,15)], sd)

## Select the features to use to create the clusters ----
# OPTION 1: Use all the numeric variables to create the clusters
AirTraffic_vars <-
  AirTraffic_removed_obs_std[, c(1,12,14,15)]
# OPTION 2: Use only the most significant variables to create the clusters
#AirTraffic_vars <-
  #AirTraffic_removed_obs_std[, c("Month",
                                       #"GEO_Region")]
```


# 5. Create the clusters using the K-Means Clustering Algorithm
This code first explores the optimal number of clusters by iteratively running K-Means with different numbers of clusters and then uses a scree plot to identify the elbow point. Finally, it selects 5 clusters as the optimal number and performs K-Means clustering with 5 clusters on the dataset.
```{r step-five-chunk}
# We start with a random guess of the number of clusters we need
set.seed(7)
kmeans_cluster <- kmeans(AirTraffic_vars, centers = 4, nstart = 20)

# We then decide the maximum number of clusters to investigate
n_clusters <- 8


# Initialize total within sum of squares error: wss
wss <- numeric(n_clusters)

set.seed(7)

# Investigate 1 to n possible clusters (where n is the maximum number of 
# clusters that we want to investigate)
for (i in 1:n_clusters) {
  # Use the K Means cluster algorithm to create each cluster
  kmeans_cluster <- kmeans(AirTraffic_vars, centers = i, nstart = 20)
  # Save the within cluster sum of squares
  wss[i] <- kmeans_cluster$tot.withinss
}

## Plot a scree plot ----
# The scree plot should help you to note when additional clusters do not make
# any significant difference (the plateau).
wss_df <- tibble(clusters = 1:n_clusters, wss = wss)

scree_plot <- ggplot(wss_df, aes(x = clusters, y = wss, group = 1)) +
  geom_point(size = 4) +
  geom_line() +
  scale_x_continuous(breaks = c(2, 4, 6, 8)) +
  xlab("Number of Clusters")

scree_plot

# We can add guides to make it easier to identify the plateau (or "elbow").
scree_plot +
  geom_hline(
    yintercept = wss,
    linetype = "dashed",
    col = c(rep("#000000", 5), "#FF0000", rep("#000000", 2))
  )

# The plateau is reached at 5 clusters.
# We therefore create the final cluster with 5 clusters
# (not the initial 4 used at the beginning of this STEP.)
k <- 5
set.seed(7)
# Build model with k clusters: kmeans_cluster
kmeans_cluster <- kmeans(AirTraffic_vars, centers = k, nstart = 20)

```


# 6. Add the cluster number as a label for each observation
This code adds cluster labels to a dataset that has undergone K-Means clustering. It then creates a scatter plot to visualize the data points, with each point color-coded according to its cluster assignment.
```{r step-six-chunk}
AirTraffic_removed_obs$cluster_id <- factor(kmeans_cluster$cluster)

## View the results by plotting scatter plots with the labelled cluster ----
ggplot(AirTraffic_removed_obs, aes(Passenger_Count, Month,
                                         color = cluster_id)) +
  geom_point(alpha = 0.5) +
  xlab("Number of Passengers") +
  ylab("Month")

```
